The Pair Distribution Function of the Inhomogeneous Hard-Sphere Fluid

### 2014-03-07 papers/pair-correlation/paper.tex
more edits on paper plus reenable this-work-mc Tested on bennet. Took 28:53.21.

At its core, thermodynamic perturbation theory---sometimes referred to
as the cluster, or high-temperature expansion---is an expansion of the
free energy in powers of a small parameter, which typically is a
pairwise attractive interaction:
\begin{align}
  F &= F_0 + F_1 + \beta F_2 + \mathcal{O}(\beta^2)
\end{align}
where the terms $F_n$ are corrections to the free energy at order $n$
in the small interaction.  The first and largest term in this
expansion is
\begin{align}
  F_1 &= \tfrac12 \iint g^{(2)}_{HS}(\rr_1,\rr_2)n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|) d\rr_1d\rr_2
\end{align}
where $g^{(2)}_{HS}(\rr_1,\rr_2)$ is the pair distribution function of
the hard-sphere reference fluid, and $\Phi(r)$ is the pair potential.
In this paper, we will introduce an approximation for the hard-sphere
pair distribution function which is suitable for use in the creation
of classical density functionals based on a cluster expansion.

@@

\section{Our approach}
%% We approximate the pair correlation with the average of two radial
%% distribution functions, evaluated at the distance between the two
%% points, that are themselves functions of the averaged pair correlation
%% at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
%% evaluated at the two points.
In the approaches for the pair distribution function mentioned above,
the radial distribution function used to approximate pair correlation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by using instead a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}.
\end{align}

We construct our function's form in terms of an expansion of powers of
$h_{\sigma} = g_{\sigma}-1$, ensuring that it approaches $g_{\sigma}$
in the contact distance limit, leaving our function with tunable
constant perameters.  We run a simple monte-carlo simulation to
calculate values of the homogeneous radial distribution function at
varying distance perameter $r$.  We then tune our function's constant
parameters by matching our function's results against the monte-carlo
data, using a least squares fit method.  During this process it is
simplest to use an analytic expression for $g_{\sigma}$ in our
function and so we use the well known and very successful Carnahan
Starling equation to calculate $g_{\sigma}$.  This is appropriate
since our published $g_{\sigma}$ approaches the Carnahan Starling
version in the homogeneous limit and it's this limit, along with the
contact limit, that we would like our constructed radial distribution
function to get right.  After this process we have an appropriate
radial distribution function that is dependent only on $r$ and
$g_{\sigma}$.

@@

The function has the form:
\begin{multline}
  g(r,g_\sigma) = 1 + h_\sigma e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + h_\sigma(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
where $\kappaO$,$\kappaI$, and $\kappaZ$ are fitted constants and
$I(g_\sigma)$ is a function of $g_\sigma$.

The finalized form of our pair correlation function is:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
\end{align}
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}

>>>>

At its core, thermodynamic perturbation theory---sometimes referred to
as the cluster, or high-temperature expansion---is an expansion of the
free energy in powers of a small parameter, which typically is a
pairwise attractive interaction:
\begin{align}
  F &= F_0 + F_1 + \beta F_2 + \mathcal{O}(\beta^2)
\end{align}
where the terms $F_n$ are corrections to the free energy at order $n$
in the small interaction.  The first and largest term in this
expansion is
\begin{align}
  F_1 &= \tfrac12 \iint
  g^{(2)}_{HS}(\rr_1,\rr_2)n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|)
  d\rr_1d\rr_2
  \label{eq:mean-field}
\end{align}
where $g^{(2)}_{HS}(\rr_1,\rr_2)$ is the pair distribution function of
the hard-sphere reference fluid, and $\Phi(r)$ is the pair potential.
In this paper, we will introduce an approximation for the hard-sphere
pair distribution function which is suitable for use in the creation
of classical density functionals based on a cluster expansion.

@@

\section{Our approach}
%% We approximate the pair correlation with the average of two radial
%% distribution functions, evaluated at the distance between the two
%% points, that are themselves functions of the averaged pair correlation
%% at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
%% evaluated at the two points.
In the approaches for the pair distribution function mentioned above,
the radial distribution function used to approximate pair correlation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by using instead a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) +
    g(|r_{12}|, g_\sigma(\rr_2))}{2}. \label{eq:g2-our-mean}
\end{align}
This approximation is constructed to reproduce the exact value for the
integral$ \int n(\rr_1)n(\rr_2)
g^{(2)}(\rr_1,\rr_2)\delta(|\rr_1-\rr_2|-\sigma)d\rr_1d\rr_2$.  This
approximation (described in Eq.~\ref{eq:g2-our-mean}) requires the
radial distribution function as a function of $r$ and $g_\sigma$.  One
approach is to compute $\eta$ from $g_\sigma$ and then find
$g(r,\eta)$ either using some approximate functional form
(e.g. Percus-Yevick) or using Monte Carlo simulation results.  In this
paper we use the latter (also for evaluating the approximations of
earlier works), and refer to this approximation (with an exact
$g(r,g_\sigma)$) as ``this work MC.''

Equation~\ref{eq:g2-our-mean} is a reasonable approximation, but
cannot be computed efficiently with single-center convolutions as
written.  In order to ease computational application of this pair
distribution function in classical density functionals, we have
developed a separable appoximation for the function $g(r,g_\sigma)$.
\begin{align}
  g(|r_{12}|, g_\sigma) &= \sum_{i} a_i(r_{12}) b_i(g_\sigma)
\end{align}
This separable approximation allows the mean field contribution to
the free energy (Eq.~\ref{eq:mean-field}) to be computed with only
single-center convolutions, which can be efficiently computed using
fast Fourier transforms.

%% We construct our function's form in terms of an expansion of powers of
%% $h_{\sigma} = g_{\sigma}-1$, ensuring that it approaches $g_{\sigma}$
%% in the contact distance limit, leaving our function with tunable
%% constant perameters.  We run a simple monte-carlo simulation to
%% calculate values of the homogeneous radial distribution function at
%% varying distance perameter $r$.  We then tune our function's constant
%% parameters by matching our function's results against the monte-carlo
%% data, using a least squares fit method.  During this process it is
%% simplest to use an analytic expression for $g_{\sigma}$ in our
%% function and so we use the well known and very successful Carnahan
%% Starling equation to calculate $g_{\sigma}$.  This is appropriate
%% since our published $g_{\sigma}$ approaches the Carnahan Starling
%% version in the homogeneous limit and it's this limit, along with the
%% contact limit, that we would like our constructed radial distribution
%% function to get right.  After this process we have an appropriate
%% radial distribution function that is dependent only on $r$ and
%% $g_{\sigma}$.

%% The function has the form:
%% \begin{multline}
%%   g(r,g_\sigma) = 1 + h_\sigma e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
%%   + h_\sigma(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
%%   + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
%% \end{multline}
%% where $\kappaO$,$\kappaI$, and $\kappaZ$ are fitted constants and
%% $I(g_\sigma)$ is a function of $g_\sigma$.

%% The finalized form of our pair correlation function is:
%% \begin{align}
%%   g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
%% \end{align}
%% \begin{figure}
%%   \centering
%%   \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
%%   %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
%%   \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
%% \end{figure}

@@
